---
title: 'Functional programming in R: Higher-Order Functions and Closures'
author: "William Hopper"
date: "`r Sys.Date()`"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(prompt=FALSE, cache = TRUE) 
```

## Last Time...
We explored "First-Class" functions in R, and how having functions be "just another object" makes the `apply` family of functions possible (`lapply`, `sapply`, `vapply` and regular old `apply`).

Today, we're going to look at a few neat tricks that you need to have up your sleeve when doing functional-style programming in R.

  - "Purely Functional" programming without R's "Syntactic Sugar"
  - Higher-order functions: When imperative style just wont do
  - Closures (a.k.a. functions, but with some data along for the ride)

## R's Sytactic Sugar
R is a functional programming language at its core, and functional programming languages are organized around (you guessed it) functions. In "pure" functional languages, *everything* is done as a function call. 

It may not look like it, but this is true in R as well. To quote John Chambers (the designer of the R language):

> "To understand computations in R, two slogans are helpful: Everything that exists is an object. Everything that     happens is a function call."

Allow me to demonstrate: What does the following code return?

```{r eval=F}
2*(-3)
```

## What? How? Why?
We got two-thirds instead of -6 because of this particularly evil snippet of code.

It redefines the `*` and `(` operators, places them in a new environment, and inserts that environment onto the search path before the `base` package, which has the real `*` and `(` operators!

```{r evil, eval = FALSE}
e <- new.env(parent = emptyenv())
e$`*` <- function(x,y) {x/y}
e$`(` <- function(x) -x
suppressMessages(attach(e))
rm(e)
```

## Everything that happens is a function call
This works precisely because the `*` and `(` operators are functions - they just have an atypical syntax that makes them more natural to use. But, we can always use them with "traditional" function syntax by back-quoting them (which disables their typical evaluation).  

```{r}
`*` # The * function 
`*`(10,2)
```

This is true for all of R's "infix" operators: `+`, `-`, `/`, `<-`, `[`, `[[`, `!`, `==`, `%in%`, `%*%`, etc. They are all functions!

## Using infix operations in functional programming
Using infix operators in functional programming contexts can be quite useful. For example, if I had a list of vectors, and wanted to remove the first element in each vector, but still keep them in the list, I could do the following:

```{r}
x <- list(c(8, 3, 2),  # list of numeric vectors
          c(4, 5, 2),
          c(3, 10, 9))
lapply(x ,`[`, -1) # -1 is passed as the argument to `[`
```

## Higher-Order Functions
Higher-Order functions are functions that take other functions as arguments, and return a vector as output

You've already seen R's most commonly used higher-order functions (`lapply` *et al.*),  but we'll look at three more general-purpose higher-order functions today: 

- `Map()`
- `do.call()`
- `Reduce()`


## Map
`Map` is a close cousin of the `*apply` functions, and can be thought of as a "multivariate" apply function.

Instead of providing each *individual* element of a *single* vector as an argument to a function, it provides *sets* of elements from *multiple* vectors as arguments to a function.

## Map
Last week, we saw that the following two expressions were equivalent:
```{r}
x <- lapply(c(8, 6), seq, from = 3)
str(x)
y <- list(seq(from=3, 8), # doing it "manually"
          seq(from=3, 6))
str(y)
```

But what if we wanted the `from` argument to vary between calls to seq? Say we wanted the sequence going to 8 to start at 3, but the sequence going to 6 to start at 1?

## Map
When you want multiple arguments to vary on each function call, that's a job for `Map`. All we have to do is provide a second vector that has the values we want to be used as the `from` arguments.

```{r}
x <- Map(seq,
         from = c(3, 1),
         to = c(8, 6))
str(x)
y <- list(seq(from=3, to=8),
          seq(from=1, to=6))
str(y)
```

## do.call
The `do.call` function is useful when the values you want to use as arguments to a function are stored inside of a list. For example, the following two expressions are equivalent:

```{r}
x <- list(x = c(50, NA, 12, 89, 61), na.rm = TRUE)
do.call(mean, x)
mean(x$x, na.rm = x$na.rm)
```

This is commonly used together with functions that take a potentially unlimited number of arguments (e.g., `rbind` and `cbind`).

## Reduce
The `Reduce` function uses a binary function (i.e., a function that takes 2 arguments) to successively combine the elements of a vector or list. 

In other words, it *reduces* a vector down from many values to a single value by repeatedly calling a function on pairs of values from that vector.

The `Reduce` function has 2 key arguments:

- `f`: The binary function
- `x`: The vector or list whose elements will be passed to `f`

## Reduce
So if we had a vector with 4 elements (`x1`, `x2`, `x3`, and `x4`), and some function `f`, `Reduce` performs the following operation internally:

```{r eval = FALSE}
Reduce(f, c(x1, x2, x3, x4))
# is the same as:
f( f( f(x1, x2), x3), x4)
```

First, it calls `f` with `x1` and `x2` as arguments. 

Then it uses the result from `f(x1, x2)` as the first argument to `f`, along with `x3` as the second argument.

Finally, the result of that operation is used as an argument to `f` again, along with `x4`, to find the final result.

## Reduce
For example, we could use the `Reduce` function to calculate a the sum of an entire numeric vector

```{r Reduce}
1 + 2 + 3 + 4 + 5 + 6 + 7 + 8 + 9 + 10
Reduce(`+`, 1:10) # don't do this in real life, its quite slow
sum(1:10)
```

A place where this function is quite useful in "real life" is joining together several data frames. Remember, the `*_join` functions from `dplyr` only join together 2 data frames, but what if you need to join together 3, or 4 or 5?

Using `Reduce(left_join, list(df1, df2, df3, df4, df5))` is a lot better than writing out all 4 joins explicitly.

## Closures
A closure is a function with access to data *not* supplied using the argument interface. In R, the accessible data (if any) are the objects defined in the environment the function is *enclosed* in.

The environment that encloses a function is the environment where it was defined. For example, if I executed the statement `f <- function(x) { x+1 }` in the console, then function `f` would be enclosed in the global environment.

```{r}
f <- function() { x+1 }
environment(f)
```

## Closures
This relationship matters because R employs a method of matching variables (i.e., `x`) to values (i.e., `10`) known as *lexical scoping*. Lexical scoping means that if R can't find the value associated with a variable in the current environment, it looks in the enclosing environment. In other words, it looks in the environment where the function was created.

So if our function `f <- function() { x+1 }` is invoked, the variable `x` is not defined in the function's local environment, and so R will look in `f`'s enclosing environment for the value of `x`.

<div class="errormsg">
```{r, error=TRUE}
f <- function() { x + 1 }
f()
```
</div>

```{r}
x <- 10 # Define x in the global env
f()
```

## Closures
This relationship becomes more interesting if a function is defined inside another function. 

Let's say we define function `g` inside the body of function `f`, and subsequently return the function object `g`. Then, function `g`'s enclosing environment is function `f`'s local "runtime" environment.

```{r}
f <- function() {
  print(environment()) # Get memory address of local environment
  g <- function() { x + 1}
  return(g)
}
g_prime <- f()
environment(g_prime)
```

## Closures
This means that the function `g` can access and use any variables defined inside `f`'s environment.

Who can predict what `g_prime()` will return?
```{r add_one, eval=FALSE}
f <- function(x) {
  g <- function() { x + 1}
  return(g)
}
g_prime <- f(10)
g_prime()
```

## Closures
Yes, `g_prime()` will return 11.

```{r,ref.label="add_one"}
```

That's because the variable `x` was assigned a value of 10 inside `f`'s local execution environment by calling `f(10)`, and because `g` has access to variables defined in its enclosing environment.

## Activity
Execute the following statements, which will write several .csv's to a temp folder and create a vector called `filenames` which holds the full path to each file.
```{r write, results='hide'}
set.seed(1)
tmp <- tempdir()
sapply(1:10, function(x) {
  write.csv(as.data.frame(matrix(rnorm(1000),100, 10,
                                 dimnames = list(NULL, letters[1:10]))
                          ),
            file = file.path(tmp, paste0("FakeSubject", x, ".csv")),
            row.names = FALSE)
  })
filenames <- list.files(path = tmp, pattern = "FakeSubject[0-9]+.csv",
                        full.names = TRUE)
```

## Activity
Combine the data in each csv into a single data frame. The catch: you may only use a single statement that must be 50 characters or less (not counting variable assignment).

Bonus question:

Which row in this data frame is the only row to have all observations less than 0?

```{r solutions, include = FALSE, eval = FALSE}
x <- do.call(rbind, lapply(filenames, read.csv))
which(Reduce(`&`, lapply(x, `<`, 0)))
which(t(apply(x, 1, function(y) all(y<0))))
```
