---
title: 'dplyr, Part 1: Basic Verbs'
author: "William Hopper"
date: "`r Sys.Date()`"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(prompt=TRUE, cache = TRUE) 
```

## What is dplyr?

The `dplyr` package is quickly becoming the de-facto standard for common tasks that involve manipulating data frames. Some of its biggest advantages are: 

- A streamlined and unified interface for tasks commonly performed together. 
- More syntactically and semantically intuitive functions than the tools in base R
- Often much faster!
- The ability to manipulate data stored in databases (e.g. MySQL, SQLite, PostgreSQL) with the same language you use for 'ordinary' data in R (i.e., data frames).

This is not to say that it replaces or makes base R syntax obsolete, but it is a great tool to have at your disposal.

## Functions == Verbs in dplyr
`dplyr` provides some basic "verbs", i.e., functions that correspond to the most common data manipulation tasks. 

This model helps build a bridge between your thoughts (i.e., what you want to do to the data, like "I want to select only observations from the first time point") into code. For example:

- "I want to sort the rows" --> `arrange()`
- "I want to select only some rows" --> `filter()`
- "I want to select only some columns" --> `select()`
- "I want to find the unique rows" --> `distinct()`
- "I want to create a new column" --> `mutate()`
- "I want to aggregate the data" --> `summarise()`

## Getting Started
We'll start our tour by installing the `dplyr` package and loading a dataset to use in our examples.

```{r installing,eval=FALSE}
install.packages('dplyr')
library(dplyr)
cropYield <- read.csv("http://wjhopper.github.io//psych640/data/crops.csv")
head(cropYield)
```

```{r realInstall, echo=FALSE, message=FALSE, warning =FALSE}
library(dplyr)
cropYield <- read.csv("../data/crops.csv")
head(cropYield)
```

The `cropYield` data frame has 99 observations of 4 variables: Field, Till,	Fert,	and Yield.

## The arrange function
The `arrange` function sorts the rows of a data frame based on the values of one or more variables. Lets use it to sort the rows of the `cropYield` data frame by fertilizer type.

`arrange` expects the first argument to be the data frame to sort, and uses any remaining arguments as the names of the variables to sort by.
```{r arrange1, R.options = list(max.print = 20)}
arrange(cropYield, Fert) # Column names are unquoted!!
```

## The arrange function
If you ask `arrange` to sort by multiple variables, the rows are sorted hierarchically according to the order of the variables listed in the function call. 

```{r arrange2, R.options = list(max.print = 20)}
arrange(cropYield, Fert, Yield)
```

## The arrange function
You can also sort by reverse alphabetical or numerically descending order by using the `desc` function inside the call to `arrange`.

```{r arrange_desc, R.options = list(max.print = 20)}
arrange(cropYield, Fert, desc(Yield)) # desc only works inside arrange
```

## The filter function
The `filter` function selects a subset of rows from the data frame based on logical tests, similar to the `[` operator.

`filter` also expects the first argument to be a data frame. Any other arguments must be expressions that return logical vectors, and `filter` returns only the rows where the logical conditions are met.

```{r filter1,R.options = list(max.print = 16)}
# Find the rows where the tilling method is ridge
filter(cropYield, Till == "Ridge")
```

## The filter function
If you list multiple conditions, they are combined with the `&` operator, meaning that a row must meet *both* conditions to be included in the output. 

Here, a row must have it's `Till` value equal to `Ridge`, and it's `Fert` value equal to `Deep` to be part of the output.
```{r filter2, R.options = list(max.print = 24)}
filter(cropYield, Till == "Ridge", Fert == "Deep")
```

## The filter function
If you want to include a row if it meets *any* given condition (a logical `OR` statement), you must combine the conditions yourself with the `|` operator.

```{r filter_OR, R.options = list(max.print = 24)}
filter(cropYield, Till == "Ridge" | Fert == "Deep") 
```

## The filter function
If you want to include rows where the value of a variable is a member of a larger subset (e.g., rows where x is 1 or 2, instead of just 1), you can use the `%in%` operator.

Here we filter out the rows where the `Till` variable has a value in the set defined by the character vector `c("Ridge","Chisel")`.
```{r filter_IN, R.options = list(max.print = 20)}
filter(cropYield, Till %in% c("Ridge", "Chisel"))
```

## A note on using `%in%`
Sometimes it makes sense to use the `%in%` operator to with a set of just one value. A good example is when you're filtering on a variable which has `NA` values. If you try to use the `==` operator to filter the values, your results will have the rows which meet this criterion, **AND** all the `NA` values! 

This occurs because R treats `NA` as meaning literally missing, thus there is no second value to compare to, and thus the result of the logical test is itself missing.

But if you use the `%in%` operator, the `NA` values will be judged "not in the set", and not included in the results.

```{r in_vs_equals}
c(NA == 1, NA %in% 1) # Logical Weirdness, Exhibit A!
```

## The select function
The `select` function allows you to pare down your data set by only keeping the columns that you specify, and also allows you to rename and reorder the selected columns on the fly.
```{r select1, R.options = list(max.print = 24)}
# Existing column name on the right of the =, new name on the left 
select(cropYield, fertilizer = Fert, bushels = Yield, Till)
```

## The select function
The `select` function even allows you to select a continuous range of columns using the start and end column names, separated with the `:` operator, just like a numeric vector!!!
```{r select2, R.options = list(max.print = 20)}
# So awesome! But, you can't rename at the same time with this syntax
select(cropYield, Till:Yield)
```

## The select function
You can even specify just the columns you *don't* want to keep, which is useful for very wide data frames where you want to discard just one or two columns
```{r select4, R.options = list(max.print = 20)}
# The - says to drop this column
select(cropYield, -Field)
```

## The rename function
You can use the `rename` function if you want to change any column names without removing any columns.
```{r rename, R.options = list(max.print = 24)}
# Existing column name on the right of the =, new name on the left 
rename(cropYield, fieldID = Field)
```

## The distinct function
The `distinct` function is similar to the `unique` function, but instead of returning the unique values in a vector, `distinct` returns only the unique rows of a data frame.

Lets use it to discover all the unique combinations of `Till` and `Fert` contained in this dataset.

```{r distinct}
distinct(select(cropYield, Till, Fert))
```


## The mutate function
You can use the mutate function to create new variables in your data frame.

Its especially useful for creating new variables based on existing ones, and changing/adding multiple columns at the same time.

```{r mutate, R.options = list(max.print = 24)}
mutate(cropYield, Field = factor(Field), yeildSq = Yield^2)
```

## The mutate function
A cool feature of the `mutate` function is that you can use new variables immediately after you create them, within the same call to mutate!!

```{r mutate2, R.options = list(max.print = 24)}
# We create  yieldRoot, and then use it in the same function call
mutate(cropYield, yeildRoot = Yield^1/2, bigYield = yeildRoot + Yield)
```

## The summarise function
The `summarise` function is useful for applying functions that reduce your variables down to a single value, similar to the idea of aggregation functions in the `dcast` function from `reshape`.

```{r summarise}
summarise(cropYield, avgYield = mean(Yield))
```

This may not seem incredibly useful on its own, but allows us to segue into what is perhaps the biggest reason to use `dplyr`.

## The Killer Feature of dplyr
Unless you are struggling to work with a remote MySQL server, or waiting hours for your code to run on data frames with millions of observations, it might not seem worth to "relearn" how to do everything you know how to do.

But `dplyr` has one feature that makes it worth every minute of learning: The ability to define persistent groups in your data frame, and apply functions to *each individual group at the same time*.

## Group-wise operations, the hard way
Consider this method of finding the average crop yield for each of the unique tilling and fertilizer groups in our dataset.

```{r groupwise_bad}
ChiselBroad <- mean(cropYield$Yield[cropYield$Till == 'Chisel'
                              & cropYield$Fert == 'Broad'])
ChiselDeep <- mean(cropYield$Yield[cropYield$Till == 'Chisel'
                             & cropYield$Fert == 'Deep'])
ChiselSurface <- mean(cropYield$Yield[cropYield$Till == 'Chisel'
                                & cropYield$Fert == 'Surface'])
MoldbrdBroad <- mean(cropYield$Yield[cropYield$Till == 'Moldbrd'
                               & cropYield$Fert == 'Broad'])
# And so on, and so forth into infinity...
```
This is verbose, tedious, prone to typos, and  doesn't scale to large datasets with many groups. A `for` loop would help, but there's an even better way...

## The group_by function
`dplyr` provides a function called `group_by` you can use to specify variables in your data frame whose values (or combinations of values) define groups of observations.

For example, we can use `group_by` to group the larger `cropYield` data frame into 9 smaller groups, one for each of the tilling and fertilizer method combinations.

```{r group_by, R.options=list("max.print" = 20)}
cropsGrouped <- group_by(cropYield, Till, Fert) # create the grouped dataset
```

## The group_by function
The `group_by` function doesn't change the external appearance of the data frame; the only noticeable change is the addition of the line `Groups: Till, Fert [9]` when the data frame is printed.
```{r group_by2, R.options=list("max.print" = 16)}
cropsGrouped # Inspect the grouped dataset
```

This line tells you which columns the groups are based on, and how many individual groups they form.

## Group-wise operations, the easy way
When you apply the `mutate` and `summarise` functions to a grouped data frame, the operations you perform inside the functions are no longer applied to the entire data frame.

Instead, they are applied to each group of observations individually, **without any change to the function call**.

This technique makes finding the mean of each individual tilling/fertilizer group incredibly simple instead of complicated and tedious.

```{r good_groupby}
cropsGrouped <- group_by(cropYield, Till, Fert) # easy peasy
grpMeans <- summarise(cropsGrouped, avgYeild = mean(Yield)) # lemon squeezy
```

## Group-wise operations, the easy way
Viola!
```{r means}
grpMeans
```

## Group-wise operations
Lets take a closer look at meta-data about our summary data frame.
```{r peeling, echo=FALSE, R.options=list("max.print" = 6)}
grpMeans
```
The `Groups` line at the top has changed: now our data frame is only grouped by `Till`, and the group size is now `[?]`.

This is because `summarize` always "peels off" the final grouping variable, because it doesn't make sense to continue grouping by the same columns. If you did, each group size would be one, because you just summarized each group down to one value.

## Group-wise operations
You can use the `n` function, with no arguments, inside calls to `summarise` and `mutate` to count the number of observations in a group.
```{r n}
summarise(cropsGrouped, grpSize = n())
```

## Ungrouping
If you want to remove the grouping you've applied to a data frame, just use the `ungroup` function to reset it back to a "normal" data frame.
```{r ungroup, R.options=list("max.print" = 12)}
ungroup(cropsGrouped) # no more "Groups" line
```
You can check a data frames grouped/ungrouped status with the `is.grouped` function.

## Activity
Use the `group_by` and `summarize` functions to find the mean and standard error of the crop yields recorded for:

  - Each field
  - Each tilling method
  - Each fertilizing method
  - Each combination of tilling and fertilizing method