---
title: "Reading and Writing Data Sets"
author: "William Hopper"
date: "`r Sys.Date()`"
---

```{r setup, include=FALSE}
library(readxl)
library(R.matlab)
knitr::opts_chunk$set(prompt=FALSE,comment= NA, cache = TRUE)
commas <- read.csv("../data/commas.csv")
data <- read.fwf("../data/fixed.txt", widths=c(1,1,-5,2),
                 col.names= c("condition", "trial", "rating"))
chicks <- read_excel("../data/datasets.xlsx", sheet = 3)
sheetnames <- excel_sheets("../data/datasets.xlsx")
chickwts <- read_excel("../data/datasets.xlsx",sheet = "chickwts")
matData <- readMat("../data/matfile.mat")
```

## File I/0
- The process of reading information from a file on your hard disk, and writing information to disk is known as **File I/O**
    - Short for file input/output. 
- Base R includes many functions implementing file i/o, so you can get your data off your disk, into R for analysis, and get the final results out of R for sharing.

## File paths
- To read from or write to a file, R needs to know where that file is located before anything can be done. 
- This is accomplished by specifying the file's location in the file system hierarchy, better known as the **file path**.
- The location of a file can be supplied in 1 of 2 ways:
    1. Using a file's *absolute* path (i.e., specifying its location relative to the beginning of the of the file system)
    2. Using a file's *relative* path (i.e., specifying its location relative to R's current working directory)
    
## Absolute Paths
- On Windows, the start of each file system is given a letter prefix; the prefix of the file system holding the Window's installation is "`C:\\`"
    - Directories are separated  with **backward** slashes
    - E.g. "`C:\Users\will`" is an absolute path
- On Mac OSX, Gnu/Linux or other unix-like systems, the start of the file system is `/`
    - Read as "root"
    - Directories are separated  with **forward** slashes
    - E.g. "`/Users/will`" is an absolute path

## The working directory
- Whenever you start R, the process that is begun "lives" in a specific directory on your filesystem. 
    - The location where it "lives" is called the **working directory**
- On Windows, the default working directory is your user's "My Documents" folder
- On OSX, the default working directory is your user's directory in the `/Users` directory.
- On Linux, the default working directory is your user's directory in the `/home` directory.

You can check your working directory with `getwd()`, and change it with `setwd()`
 
## Relative Paths
So, if you know R's working directory *a priori*, you can specify the path to a file relative to the directory. 

Lets contrast the two methods. First we read the file 'commas.csv' using it's full path
```{r badexamples,eval=FALSE}
read.csv("/home/will/data/commas.csv")
```

But, our current working directory already is "/home/will"
```{r badexamples2, eval=FALSE}
getwd()
```

```{r badexamples2_2, eval=TRUE,echo=FALSE}
print("/home/will")
```

So we can remove the "/home/will" portion of the path, and use the remaining relative portion to read in the file.
```{r badexamples3,eval=FALSE}
read.csv("data/commas.csv")
```

## Special Path Characters

There are two "special characters" to be aware of when dealing with file paths. Think of them as variables that change values when you change working directories.

1. `..` : Short-hand for "one directory up" from the current working direcoty in the file system heirarchy
2. `.` : Short-hand for "the current working directory".

Current working directory | Command | New working directory
--- |  ------- | ------
`/home/will/data/` | `setwd("..")` | `/home/will`
`/home/will/data/` | `setwd("../..")` | `/home`
`/home/will/data/` | `setwd(".")` | `/home/will/data`
`/` | `setwd("..")` | `/`

## The Structure of (Good) Data
It's worth beginning with an outline of a well-formatted data set. 

- The data is represented in a rectangular structure (table with rows and columns)
- Each column represents a specific variable, with a header signifying the name of this variable
- Each row is represents an observation
- Avoids names or values with blank spaces
- Avoids using names that contain symbols such as :, ;, ?, $, %, ^, &, *, (, ), -, #, ?, < , >, /, |, [, ], { and } 
 - Any missing values in your data set are indicated with `NA`


## Plain Text Formats
- Plain text files consist only of sequences of characters codes, including spaces, tabs, new lines and delimiters.
- They have no styling associated with them (e.g. no italics or bolding, no images)
- Plain text examples: HTML, .txt files, .R files
- NOT plain text: Word documents, Excel documents

## Comma Separated Value (.csv)
In a csv file, the columns in each row are separated by commas, naturally. Below is an example of plain-text formatted as a 
.csv file:
```{r showCommas,echo=FALSE, prompt=FALSE}
write.csv(commas,file="",quote=FALSE, row.names=FALSE)
```
[Download this file](../data/commas.csv)

We can import this file into a data frame in R with the `read.csv()` function

## Commonly used `read.csv` arguments

Arg | Controls | Default Value
--- |  ------- | ------
file | Path to where file is stored | None, `file` must be specified
header | Treat the first row as column headers? | `TRUE` 
sep | What character delimits the columns within a row? | `,`
dec | What character is used for the decimal place? | `.`
quote | What character(s) is used as quotation marks | `"` or `'` 
row.names | names for the rows | None, must specify a character vector
col.names | names for the columns | None, must specify a character vector

## Commonly values for `read.csv` arguments
- Specifying the `sep` argument allows you to read in files where the columns are separated by non-standard delimiters
    - e.g. tabs (use "`\t`"), spaces (use `" "`), `_`, `;`, `:`, or `.`
- Specifying the `dec` argument allows you to read in data sets were the decimal place is denoted with `,`. 
 - If the csv file *doesn't* have a row of column headers, use `header=FALSE` and supply a character vector for the `col.names` argument to give names to the variables in the resulting data frame you create.
    - If the csv file *does* have a row of column headers, you can still override them with  the `col.names` argument.
 
## Basic `read.csv()` example
```{r readCSVexample1, eval=FALSE}
commas <- read.csv("data/commas.csv",sep=",",header=TRUE)
```
```{r readCSVexample2, eval=TRUE, echo=FALSE}
print(commas)
```
```{r readCSVexample3, eval=FALSE}
read.csv("data/commas.csv",sep=",",header=TRUE,
         col.names=c("Cond","Tr","Rat")
```
```{r readCSVexample4, eval=TRUE, echo=FALSE}
setNames(object = commas, nm = c("Cond","Tr","Rat"))
```


## Interactive File Selection
R allows you to select which file to import using the default file explorer tool from your OS. 

Instead of specifying the path to the file you want as a character vector, you instead pass the `file.choose()` function as the first argument.

```{r choosingfiles,eval=FALSE}
data <- read.csv(file=file.choose())
```

This should pop up a file selection dialog, like when adding an email attachment.

## Exporting Data Structures as a CSV
- You can write the contents of a matrix or data frame to a plain text file, such as a csv
    - Use the `write.csv()` function

Arg | Controls | Default Value
--- |  ------- | ------
file | Path to where file should be created (including file extension)| None, `file` must be specified
append | If the file already exists, should we add to it? | `FALSE` 
quote | Should characters or factors be printed with quotes | `TRUE` 
row.names | Should row names be written? | `TRUE`

## Exporting Data Structures as a CSV
Its often useful to set `quotes` and `row.names` to `FALSE` when writing a csv.

Compare the defaults: 
```{r defaultWrite,prompt=FALSE, comment=""}
write.csv(commas, file = "")
```

to this:
```{r optimalWrite,prompt=FALSE, comment=""}
write.csv(commas, file="", quote=FALSE, row.names=FALSE)
```

## Reading in fixed-width data
Sometimes you may encounter data where the columns in each row are not separated by a delimited, but by a fixed number of character spaces.

The previous file, which was separated by commas, has been changed to a fixed-width format below.

<pre class="plain">
a1xxxxx3 
b2xxxxx1 
c3xxxxx11
</pre>
<a href="../data/fixed.txt" download> Download this file </a>

We can see that the first column takes of 1 space, the second column takes up 1 space, and that the third column takes up 2 spaces.

But, the second and third column have nonsense between them, the `xxxxx` you see.

## Reading in fixed-width data
We can read this into R with the `read.fwf()` function, specifying the file path, the width of each column, and the width of things to ignore. 

```{r readinfixed1, eval = FALSE}
data <- read.fwf("data/fixed.txt", widths=c(1,1,-5,2),
                 col.names= c("condition", "trial", "rating"))
print(data)
```
```{r readinfixed2, eval = TRUE, echo=FALSE}
print(data)
```

The positive values in the `width` vector specify the number of characters to attribute to each column (i.e., 1 for the first, 1 for the second, 2 for the third).

The negative values tell R to skip characters. Here, -5 in the third position means to skip 5 characters before starting the third column (i.e. skip the x's).

## Reading data from Excel
The simplest way to read in data from Excel is to use Excel to save it as a CSV, and then use `read.csv()` =)

But otherwise he easiest way to read in data from .xls and .xlsx files is using the `readxl` package. Get it with `install.packages('readxl')`.

```{r readxl,eval=FALSE}
install.packages("readxl")
library(readxl)
```

## Reading data from Excel
The main function in the package is `read_excel()`

Arg | Controls | Default Value
--- |  ------- | ------
file | Path to where file is stored (including file extension)| None, `file` must be specified
sheet| Which page should be read in (can use name or number) | `1` 
col_names | TRUE/FALSE to use first row as column names, or a vector of name | `TRUE` 
skip| Number of rows to skip before starting import |`0`
na| Value to use for missing data | `" "`

## Reading data from Excel: Examples
First, [download this xlsx file](../data/datasets.xlsx) containing 4 "famous" R data sets (iris, mtcars, chickwts and quakes).

We can read in the `chickwts` data set (the 3rd page of the excel file) by number

```{r xlsx, eval = FALSE}
chicks <- read_excel("data/datasets.xlsx", sheet = 3)
head(chicks)
```
```{r xlsx2, eval = TRUE, echo = FALSE}
head(chicks)
```

## Reading data from Excel: Examples
Or, we can read it in by its sheet name, aided in our effort by the `excel_sheets()` function, which lists out the names of all the sheets in the workbook.

```{r xlsx3, eval=FALSE}
excel_sheets("data/datasets.xlsx") # Mostly for interactive use
```
```{r xlsx4, echo=FALSE}
print(sheetnames)
```
```{r xlsx5, eval=FALSE}
read_excel("data/datasets.xlsx", sheet = "chickwts")
```
```{r xlsx6, echo=FALSE}
head(chickwts)
```

## Reading Spreadsheets: The quick and dirty way
If we just want a one-off solution for getting the some data out of an Excel workbook, and into R, with minimal thinking, we can use the system clipboard! 

On Windows, there is a base R solution: Highlight the cells you want, copy them to the clipboard (i.e., Ctrl + c), and then in R, call `read.table("clipboard")`. 

On other platforms, use the `read.clipboard()` function from the `psych` package. I find it slightly tempermental, but still useful. 

## Other Excel-lent packages
If you want to write xlsx files to disk, or do more complex tasks utilizing Excel as a back end (e.g. writing a vector to a specific column in an Excel sheet), take a look at the `xlsx` package, and the `XLconnect` package.

Both are cross-platform, have their own strengths/weaknesses, and don't require an Excel installation.

Avoid using the outdated `RODBC`, `xlsReadWrite` and `gdata` packages for accessing Excel files. 

## Reading SPSS Files
We can use the `read_sav()` function from the `haven` packages to import .sav files from SPSS.
```{r spss_NOTRUN,eval=FALSE,prompt=FALSE}
install.packages("haven")
library(haven)
spss <- read_sav("data/example.sav") # spss values = factors?
```

If you have labeled variables in the .sav dataset, the variable labels will stored in the "label" attribute of each variable.

## Reading in SAS files
If you have SAS data files you want to import into R, the `haven` package also includes a `read_sas` function.

```{r sasexamples,eval=FALSE}
mySASData <- read_sas("data/randomSASfile.sas7bdat")
```

.sas7bdat = __SAS 7 B__ inary __DAT__ a

## Reading in MATLAB files
MATLAB supports saving variables in a proprietary format, called .mat files. We can read them into R using the R.matlab package's `readmat()` file. 

Using the example .mat file [here](../data/matfile.mat)
```{r mfiles, message=FALSE, warning=FALSE, eval=FALSE}
install.packages("R.matlab")
library(R.matlab)
matData <- readMat("data/matfile.mat")
```

## Reading in MATLAB files
To suppress printing of many extraneous details about the .mat file, set it's attributes to `NULL` in R
```{r mfiles2}
attributes(matData) <- NULL 
print(matData)
```

## Reading Data from your brain!
OK, so there isn't an R package for reading your mind. But, you can manually input data into an R data frame!

Call `edit(data.frame())` and watch the magic!

I'll do this myself, and walk through it with you. 

## Other Data Types
This is far from an exhaustive list of the data formats the R can read in. If you have data as XML, JSON, HTML tables or Google Spreadsheets, there are R packages/function for reading them into R's data structures. 

If your data lives in a database, like MySQL, PostgreSQL, MonetDB, MongoDB, etc., etc., there are also many packages which have 2-way interaction between R and these databases.

A good jumping off point for seeing whats out there in terms of file-reading options are these two tutorials [here](http://blog.datacamp.com/r-data-import-tutorial/) and [here](http://www.r-bloggers.com/importing-data-into-r-part-two/)
