---
title: "Lab Final"
author: "William Hopper"
date: "`r Sys.Date()`"
output:
  ioslides_presentation:
    mathjax: default
---

## Question 1
Write a program that prints the numbers from 1 to 100, in ascending order. 

But, if the number to be printed is divisible by 3, print "Fizz" instead of the number.

If the number to be printed is divisible by 5, print "Buzz" instead of the number.

And if the number to be printed is divisible by both 3 and 5, print "FizzBuzz" instead of the number.

## Question 2
Complete the following code snippet to write a program that prints the numbers from 100 to 1, in descending order.

```{r eval=FALSE}
for (i in 1:100)
```

As a constraint, you cannot write any code to precede this snippet, and you may not use a second `for` loop.

# Solutions

## FizzBuzz

```{r fizzbuzz_solution,eval=FALSE}
for (i in 1:100) {
  
  divisible_by_3 <- i %% 3 == 0
  divisible_by_5 <- i %% 5 == 0
  
  if (divisible_by_3 && divisible_by_5) {
    print("FizzBuzz")
    
  } else if (!divisible_by_3 && divisible_by_5) {
    print("Buzz")
    
  } else if (divisible_by_3 && !divisible_by_5) {
    print("Fizz")
    
  } else {
    print(i)
  }
}
```

## Reverse-printing

```{r reverse_print, eval=FALSE}
for (i in 1:100) {
  print(101 - i) 
}
```

We need to print 100 first (a large number) but our iterator (i) starts out small. If we subtract a small number from a large number, we'll still have a large number. But what large number should we subtract from?

$$
x - i = 100
$$

We know $i$ starts at 1

$$
x - 1 = 100
$$

Simple rearranging shows

$$
x = 101
$$