---
title: "Manipulating Data Part 1: The Data Awakens"
author: "William Hopper"
date: "`r Sys.Date()`"
output:
  ioslides_presentation:
    css: ["../theme/css/custom.css", "../theme/css/ManipulatingData.css"]
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(prompt=TRUE, cache = TRUE) 
```


## Indexing 

All throughout your data analysis pipeline, you will face the need to take smaller chunks out of a larger data structure. Sometimes you will need to change the data that is stored in your structure, or use it as input to another function, or perhaps you need to plot it. 

The task of slicing a smaller chunk out of a larger data structure is called **indexing** in R (sometimes called **subsetting** depending on context), and is performed using the square bracket characters `[]`. 

## Indexing 'Ingredients'

Indexing a data structure in R requires 4 ingredients. 

1. An R object that supports indexing 
    + e.g. A data frame, matrix, vector or list
2. An opening square bracket `[`
3. 1 (or more) vectors which indicate which values from the larger data structure should be pulled out. 
4. a closing square bracket `]`

Arrange the ingredients in an R expression like so:

`DataStructure[IndexVector]`

We will focus on learning what can go inside the square brackets for different types of R data structures

## Indexing Vectors

Broadly, there are 2 types of indexing vectors that are useful inside the square brackets. 

1. Numeric Vectors
    + e.g. `c(1,4,5,6,10)`
2. Logical Vectors 
    + e.g. `c(TRUE,FALSE,FALSE,TRUE)`
    
We'll start with numeric indexing vectors to get a feel for the general procedure, and move up to logical indexing.

## Numeric Index Vectors 
We'll start by slicing smaller chunks out of a larger vector. Here, the numeric vectors inside the brackets tell R the *positions* of the elements we wish to extract.
```{r large_and_smaller}
alphabet <- c("a","b","c","d","e","f","g","h","i","j","k","l","m","n","o",
              "p","q","r","s","t","u","v","w","x","y","z") 
alphabet[c(1,26)] # Extract First and 26th element
alphabet[10:20] # Extract tenth through 20th
alphabet[seq(from=1,to=26,by=2)] # Extract every other element
```

## Common Errors with Numeric Indices
The most common mistake is including a value in your indexing vector which is greater than the length of the vector you are subsetting
```{r outofbounds, error=TRUE}
alphabet[100] # there are not 100 letters in the alphabet
```
The `NA` means the value is missing. This is commonly referred to as an "index out of bounds" error.

Another common mistake is forgetting to concatenate the values you want to use for the indexing vector (i.e. forgetting the `c()` function).
<div class = "errormsg">
```{r no-c, error=TRUE}
alphabet[1,5,10]
```
</div>

## Indexing Tricks

Instead of creating a vector of values you *do* want to pick out, it may be easier to come up with a vector of ones you *don't* want. We can use negative number's to specify which vector elements we don't want. 

```{r negatives}
alphabet[c(-1,-26)] # Same as alphabet[2:24]
alphabet[-1:-10]
```
Indexing with positive vectors is usually preferred, as the intent of the code is more clear, but sometimes this is easier and more clear (e.g. when dropping the first or last value)

## Activity
Execute the following code and look at the values in `months`:
```{r example1, prompt=FALSE}
months <- c("January", "February", "March", "April", "May", "June", "July",
            "August", "September", "October", "November", "December")
months
```

Now, do the following:

1. Index `months` to pull out `"February"` and `"March"`
2. Index `months` to pull out every third month

## Logical Indexing

When performing logical indexing, you supply a vector specifying whether to extract a specific element (with a `TRUE`) or to *not* extract a specific element (with a `FALSE`).

Let's revisit the example of selecting the first and last elements of the alphabet vector: We make a vector of logicals and stick it in the square brackets after your vector.

```{r bad_logical_indexing}
alphabet[c(1,26)]
alphabet[c(TRUE, FALSE, FALSE, FALSE, FALSE, FALSE, FALSE, FALSE, FALSE,
           FALSE, FALSE, FALSE, FALSE, FALSE, FALSE, FALSE, FALSE, FALSE,
           FALSE, FALSE, FALSE, FALSE, FALSE, FALSE, FALSE, TRUE)]
```

## Logical Indexing

But this specific example is not a good use case for logical vectors. Why?

1. Longer Code: length of the logical vector must match the length of the object its subsetting.
2. Duplicating work: If you already know the position of the elements you want, just put them into a vector and you're done!

The logical vector's utility comes into play when you *don't* know the numeric positions of the elements you are interested in.

## Logical Indexing

"But wait" you say: "If we don't know where they are already, how are we going to find them?". This brings us to **logical testing** and **relational operators**

Relational operators are R expressions that can be used to test whether or not some value meets a condition or not. If it meets your test's conditions, the test returns `TRUE`. If it does not, it returns `FALSE`. 

You already know lots of relational operators. The equal to, greater than and less than expressions from 3rd grade math are all relational operators!

## Relational Operators Table

Comparison | Expression | Example
---------- | ---------- | -------
Less Than | `<` | `5 < 10`, `5 < 1`
Less Than or Equal To | `<=` | `5 <= 5`, `5 <= 1`
Greater Than | `>` | `10 > 5`, `5 > 5`
Greater Than or Equal To | `>=` | `10 >= 10`, `10 >= 12`
Equal To | `==` | `5 == 6`, `5 == 5`
Not equal to | `!=` | `5 != 6`, `5 != 5`

## Relational Operators & Indexing
What makes relational operators useful is that they can be applied to all the elements of a data structure simultaneously. 
```{r sometest}
x <- 2:11
x <= 5 # Apply the less than or equals test
```
As you can see, values that meet the criteria (<= 5) return as `TRUE`. 
```{r sometest_pt2}
x[x <= 5] # Index vector x with the results of the test. 
```
When this logical vector is used to index the vector `x`, only the elements where the logical vector has value `TRUE` are returned.

## Relational Operators & Indexing
We index character vectors using the `==` and `!=` operators, but not the greater/less than operators. Quantity makes no sense for characters!
```{r charTesting}
months == "June" # The sixth element is TRUE
months[months == "June"]
months[months != "July"]
```

## Other Useful Tests: `is.na()`
Unfortunately, we often have to deal with missing observations in real world data sets. R codes missing data as `NA` (or sometimes `NaN`). We can use the `is.na()` function to find any missing values in a vector. 
```{r isna}
missingno <- c(10,NA,1,4,2,NA,NA,99,NaN, NA)
is.na(missingno)
missingno[!is.na(missingno)] # Select things that are the opposite of missing
```

## Other Useful Tests: modulus division
An important arithmetic operator that was not mentioned last week
is the modulus operator `%%`, which gives us the remainder of division.  

For example: `5 %% 3` is `2`.

```{r lisa}
lisa <- c(34, 509, 63, 187, 998, 78, 3330)
lisa %% 17 == 0
lisa[lisa %% 17 == 0]
```

This returns all the values of the vector called `lisa` that are multiples of 17, since multiples of 17 will have a remainder of zero.

## Tips and Tricks
A useful function to know is `which()`. When used on a logical vector, it will return to your the position indices of the vector's `TRUE` element. It is useful when you want to know **where** in the vector your matches occur. 
```{r which}
lisa %% 17 == 0
which(lisa %% 17 == 0)
```

## Activity 
Write an expression using logical testing and indexing that, when applied to the vector `dummy`, returns the same output as the numeric indexing example shown below. 
```{r indexingActivity}
dummy <- 17:23
dummy[5:7]
```

```{r indexingActivity_answer,include=FALSE}
dummy[dummy >= 21]
```

## Testing Multiple Conditions 
Sometimes you need to select elements based on multiple conditions. For example, you might want to select only those values that are less than 4 standard deviations above or below the mean. 



In R, we can select elements based on multiple conditions by combining individual logical tests together using *logical operators*. The logical operations we have at our disposal are: 

- AND
    - Each element must meet **all**  conditions to return `TRUE`
- OR
    - Each element must meet **at least one** conditions to return `TRUE`
- Negation
    - Reverse the current logical (e.g., `TRUE` becomes `FALSE`)

## Logical Operators 
Logic | Expression | Example | Result
----- | ---------- | ------- | ------
Elementwise AND | `&` | `c(1,3) > 0 & c(1,3) <=2` |`r c(1,3) > 0 & c(1,3) <=2` 
Scalar AND | `&&` | `3 > 0 && 3 <=2` | `r 3 > 0 && 3 <=2`
Elementwise OR | **`|`** | `c(1,3) > 0 | c(1,3) <=2` | `r c(1,3) > 0 | c(1,3) <=2` 
Scalar OR | **`||`** |` 3 > 0 || 3 <=2` | `r 3 > 0 || 3 <=2`
Negate | `!` | `!(3 > 0)` | `r !(3 > 0)`


The elementwise operators test all their arguments (i.e., they test all the pairs elements of the logical vectors supplied) and return a vector the same length as the input.

The scalar operators only test the *first pair* of elements from their input, regardless of length, and return a single `TRUE` or `FALSE`. Thus, scalar operators should only be used for comparing a single value to another single value.

## Indexing with Multiple Conditions 
Let's say we wanted to select the elements in `lisa` which were less than 500 or greater than 1000.

```{r show_lisa_again}
lisa
```

To do this, we need to:

1. Test each element of `lisa` to see if it is less then 500
    - `lisa < 500`
2. Test each element of `lisa` to see if it is greater than 1000
    - `lisa > 1000`
3. Combine the results of both tests together into a single test
    - ???

    
## Indexing with Multiple Conditions {.conditionals}
Our goal of the test is to have it return a `TRUE` for each element if it passed the "less than 500" test **OR** if it passed the "greater than 1000" test.

Expression | Element 1 | Element 2 | Element 3 | Element 4 | Element 5 | Element 6 | Element 7
------------- | --------- | --------- | --------- | --------- | --------- | --------- | ---------
`lisa` |     34    |   509     |    63     |    187    |    998    |     78    |    3330
`lisa < 500` | `TRUE` | `FALSE` | `TRUE` | `TRUE` | `FALSE` | `TRUE` | `FALSE`
`lisa > 1000` | `FALSE` | `FALSE` | `FALSE` | `FALSE` | `FALSE` | `FALSE` | `TRUE`
Pass Either Test? | <span style="color: Green">&#10003;</span> | 	 <span style="color: Red">&#10005; </span> | <span style="color: Green"> &#10003; </span> | <span style="color: Green"> &#10003; </span>| 	 <span style="color: Red"> &#10005; </span> | <span style="color: Green"> &#10003; </span>| <span style="color: Green"> &#10003; </span>


## Indexing with Multiple Conditions 
Our goal of the test is to have it return a `TRUE` for each element if it passed the "less than 500"" test **OR** if it passed the "greater than 1000"" test.

We can do that by combining the two tests with the `|` logical operator

Expression | Element 1 | Element 2 | Element 3 | Element 4 | Element 5 | Element 6 | Element 7
------------- | --------- | --------- | --------- | --------- | --------- | --------- | ---------
`lisa` |     34    |   509     |    63     |    187    |    998    |     78    |    3330
`lisa < 500` | `TRUE` | `FALSE` | `TRUE` | `TRUE` | `FALSE` | `TRUE` | `FALSE`
`lisa > 1000` | `FALSE` | `FALSE` | `FALSE` | `FALSE` | `FALSE` | `FALSE` | `TRUE`
`lisa < 500 | lisa > 1000` | `TRUE` | `FALSE` | `TRUE` | `TRUE` | `FALSE` | `TRUE` | `TRUE` 

<br>
```{r lisa_or}
lisa[lisa < 500 | lisa > 1000]
```

## Indexing with Multiple Conditions 
Now let's say we wanted to select the elements in `lisa` which were less than 100 and greater than or equal to 10.

To do this, we need to:

> - Test each element of `lisa` to see if it is less then 100
  >     * `lisa < 500`
> - Test each element of `lisa` to see if it is greater than 1000
  >     * `lisa >= 50`
> - Combine the results of both tests together into a single test
  >     * `&`

## Indexing with Multiple Conditions 
Our goal of the test is to have it return a `TRUE` for each element if it passed the "less than 500" test **AND** if it passed the "greater than 50" test.

Expression | Element 1 | Element 2 | Element 3 | Element 4 | Element 5 | Element 6 | Element 7
------------- | --------- | --------- | --------- | --------- | --------- | --------- | ---------
`lisa` |     34    |   509     |    63     |    187    |    998    |     78    |    3330
`lisa < 500` | `TRUE` | `FALSE` | `TRUE` | `TRUE` | `FALSE` | `TRUE` | `FALSE`
`lisa >= 50` | `FALSE` | `TRUE` | `TRUE` | `TRUE` | `TRUE` | `TRUE` | `TRUE`
`lisa < 500 & lisa >= 50` | `FALSE` | `FALSE` | `TRUE` | `TRUE` | `FALSE` | `TRUE` | `FALSE`

<br>
```{r lisa_and}
lisa[lisa < 500 & lisa >= 50]
```

## Indexing with Multiple Conditions 
```{r lisa_nothing}
lisa[lisa < 500 & lisa >= 998]
```
When you index any structure in R with a vector of all `FALSE`s, you get back `numeric(0)` which means "nothing to see here!"

So why does this test and subset return nothing? Because no number can be less than 500 and greater than 998, so every test comes back `FALSE`, and all elements are ignored in the subset.

## Exercise
1. Use R to find all the numbers between 1 and 10,000 that are multiples of 2 or multiples of 3. How many are there?
```{r exercise_ans1,include=FALSE}
numbers <- 1:10000
multiples <- numbers %% 2 ==0 | numbers %% 3 ==0
sum(multiples)
```
2. Use R to find all the numbers between 1 and 10,000 that are multiples of both 2 and 3. How many are there?
```{r exercise_ans2,include=FALSE}
multiples <- numbers %% 2 ==0 & numbers %% 3 ==0
sum(multiples)
```