---
title: "Control-Flow Statements"
author: "William Hopper"
date: "`r Sys.Date()`"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(prompt=TRUE, cache = TRUE) 
library(DiagrammeR)
```
 
 
```{r diagrams, include=F}
linear_flow <- grViz("
digraph linear {

graph [rankdir = LR]

node [shape = circle,
      fixedsize = true,
      width = 0.9]

A; B; C; D;
A ->B; B -> C; C->D; 
}",height = 150)

loop_flow <- grViz("
digraph linear {

graph [rankdir = LR]

node [shape = circle,
      fixedsize = true,
      width = 1,
      fontsize = 18]

A; B; C; D; E;
A ->B; B -> C; C->D; D->E;
C->C [label='repeat step C x times' tailport='e' headport = 'w' fontsize = 18];
}", height = 200)

branch_flow <- grViz("
digraph linear {

graph [rankdir = LR]

node [shape = circle,
      fixedsize = true,
      width = 1,
      fontsize = 18]

A; B; C; D; E;
A->B
B->C [label='x > y' fontsize = 18]
B->D [label='x < y' fontsize = 18]
D->E
C->E
}", height = 225)

```

## Go with the "Flow"
- Any program will be designed to perform some set of computational operations, for example:
    - Calculate summary statistics
    - Plot raw data
    - Do a t-test
- "Control Flow" describes the order in which specific operations take place
- A key objective when writing your code is to make your program's control flow as simple as possible while minimizing redundancy.

## Linear Program Flow 
- So far, the control flow of your code has been solely determined by the line number of each expression
    - "Code rolls downhill"
```{r linear_example, prompt=FALSE}
group1 <- sleep[sleep$group==1,] # Separate group 1 
group2 <- sleep[sleep$group==2,] # Separate group 2
x1 <- mean(group1$extra) # Calculate group 1 mean
x2 <- mean(group2$extra) # Calculate group 2 means
sem <- sqrt(.5*(var(group1$extra) + var(group2$extra)))*sqrt(2/10) # Calculate SEM
t <- (x1-x2)/sem # Find the t score
```

## Linear Program Flow 
- The control flow diagram for this script is shown below
    - A. Separate groups
    - B. Calculate the mean
    - C. Calculate SEM
    - D. Find the t score
    
```{r linear,echo=F}
linear_flow
```

## Altering the Flow
- R has expressions which can alter this linear flow. Specifically, we can:
    - Execute a set of statements *only if* some condition is met
        - `if else` statements
    - Execute a set of statements *zero or more times, until some condition is met*
        - `for` and `while` statements
    - Stop the program, preventing any further execution
        - `stop` statements

## if else syntax
An `if else` statement allows us to conditionally evaluate specific statements. The general syntax is: 

```{r ifelse_template, prompt = F, eval=F}
if (condition) {
  expressionA
} else {
  expressionB  
}
```
- `condition` is an R expression that evaulates to a single logical value (i.e. a single `TRUE` or `FALSE` value)
    - This expression **must** be in parenthesis
- `expressionA` and `expressionB` are two different blocks of R code 
    - These expressions should be contained in curly brackets

## if else control flow
```{r printBranching,echo=FALSE}
branch_flow
```

- if x < y, take the path through C
- if y > x, take the path through D
    
## if else
Lets take a look at some simple examples of conditional branching. Here we'll use an `if else` statement to print out which of two random intergers is larger. 
```{r  ifelse_ex1,prompt=FALSE,eval=TRUE}
x <- sample(1:50000,1) # random value from 1 to 50,000
y <- sample(1:50000,1)
if (x < y) {
  print("Y Wins!")
} else {
  print("X Wins!")
}
```

## if else
But we forgot a corner case - what if y equals x? We can add on additional if into this same statement like so:
```{r  ifelse_ex2,prompt=FALSE,eval=TRUE}
x <- sample(1:50000,1) # random value from 1 to 50,000
y <- sample(1:50000,1)
if (x < y) {
  print("Y Wins!")
} else if (y < x) {
  print("X Wins!")
} else {
  print("zomg X equals Y!!")
}
```

We can have as many `else if` statements chained together as we want, but only one `else`.

## if else
We can also use just the `if` part of the statement.

Here, we sample a random integer for `x`, and if it comes out negative, we flip the sign with the `abs()` function.
```{r ifelse_ex3,prompt=FALSE,eval=TRUE}
x <- sample(-10:10,1)
if (x < 0) { 
  x <- abs(x)
}
```

## Activity

Sample 2 random values from a standard normal distribution, using `rnorm()`. Call them x and y.

Then, write an if-else that does the following: 

- If they are both greater than the mean of that normal distribution, print "x and y are above the mean".
- If they are both less than the mean of that normal distribution, print "x and y are below the mean".
- If x and y fall on either side of the mean, print "x and y straddle the mean"

```{r activity_solution, include=FALSE}
rv <- rnorm(2)
if (rv[1] && rv[2] > 0) {
  print("x and y are above the mean")
} else if (rv[1] && rv[2] < 0) {
  print("x and y are below the mean")
} else {
  print("x and y straddle the mean")
}
```

## Loops 
Loops are a class of control flow that allow you to repeat the same operations over and over again, without explicitly writing out those operations over and over again in the code. 

- `for` loops repeat your operations for an explicit number of repetitions
- `while` loops repeat your operations an indeterminite number of times, stopping only when a prespecified condition occurs (whenever that might be)
```{r looping,echo=F}
loop_flow
```

## for loops 
At the heart of a `for` loop is the R object which serves as a counter. This counter variable determines the number of iterations your `for` loop should have. 

It is common to refer to the counter variable as the variable you "loop over".

In this example, we use a for loop to print out the integers 1 through 3, one at a time.
```{r basicForloop, prompt=FALSE}
for (i in 1:3) {
  print(i)
}
```

## for loops 
But the counter variable doesn't have to be sequential integers enumerating which iteration of the loop you are on. It can be any vector, and i will sequentially take on each value in the vector.
```{r basicForloop2, prompt=FALSE}
for (i in c("A","B","C",1,2,3)) {
  print(i)
}
```

## for loops 
```{r prompt=FALSE,eval=FALSE,ref.label='basicForloop'}
```

- Note 2 important things that happened:
    1. The `print()` statement was executed exactly 6 times, the same number of times as the length of `c("A","B","C",1,2,3)`
    2. The value of `i` changed on every iteration of the loop, as evidenced by the fact that `print(i)` spit out a different value every time
    
## for loops s
- For loops are most often used for their "side effects"
    - This means you loop over one object (eg. `i = 1:10`), but use the value of `i` inside the loop as a subscript to index into other data structures
    - This technique can be used to modify objects in the global environment, e.g., adding new values to an existing data frame
    
## for loops
To demonstrate this indexing, here we: 

1. Create 4 element character vector
2. Create another 4 element vector of numeric indices
3. Loop over those indices, and use the iterator to index into the character vector and print it's value. 
```{r easyforindexing}
letters <- c("A","B","C","D")
indices <- 1:4
for (x in indices) { print(letters[x])}
```
 
## Group Means Example
A common problem to solve in R is calculating summary statistics for each group in some data set. We're going to tackle this problem in the `InsectSprays` data set. 
```{r headInsectSprays}
head(InsectSprays,5)
```

First we'll walk through how we would accomplish this process with a purely linear control flow, and then we'll see how to approach the problem with a loop. 

## Group means: the hard way
First we need find the unique sprays used, and create a "placeholder" data frame which will end up storing the mean for each spray.
```{r hardway1}
sprays <- unique(InsectSprays$spray)
sprayMeans <- data.frame(spray = sprays, meanCount = NA)
sprayMeans
```

## Group means: the hard way
Then we select the first spray, and use `$` and `==` to extract the observed counts for that spray group from the `InsectSprays` data frame.
```{r hardway2}
sprays <- unique(InsectSprays$spray)
sprayMeans <- data.frame(spray = sprays, meanCount = NA)
i = sprays[1]
counts <- InsectSprays$count==i 
```

## Group means: the hard way
Then we average those extacted counts, subset our place holder data frame, and fill in the count column for this specific spray with the mean value.
```{r hardway3}
sprays <- unique(InsectSprays$spray)
sprayMeans <- data.frame(spray = sprays, meanCount = NA)
i = sprays[1]
counts <- InsectSprays$count[InsectSprays$spray==i]
m <- mean(counts)
sprayMeans[sprayMeans$spray==i,'meanCount'] <- m
sprayMeans
```

## Group means: the hard way
```{r spraymean,echo=FALSE}
sprayMeans
```
OK, one group down, 5 to go! We will repeat the same thing over and over and over again for each group. If only there was a way to get R to repeat the same operations over and over and over again....

Oh wait, `for` loops do just that!

## Group means: `for` loop
We start the same way as before, finding the unique sprays used, and creating a "placeholder" data frame to store results.
```{r sideeffects_step1, eval=F,prompt=FALSE}
sprays <- unique(InsectSprays$spray)
sprayMeans <- data.frame(spray = sprays, meanCount = NA)
```

## Group means: `for` loop   
But this time we use a for loop, looping over each unique spray. Inside the loop, we:

1. Subset out the observed counts for a particular spray
2. Find the mean of those counts, and fill that mean into the placeholder data frame
```{r sideeffects_step2,prompt=FALSE,eval=TRUE}
sprays <- unique(InsectSprays$spray)
sprayMeans <- data.frame(spray = sprays, meanCount = NA)

for (i in sprays) {
  counts <- InsectSprays$count[InsectSprays$spray==i]
  sprayMeans[sprayMeans$spray==i,'meanCount'] <- mean(counts)
}
```

## Group means: `for` loop
The result is a data frame the holds the mean insect count for each unique spray used, with no code duplication
```{r sideeffects_step3, eval=TRUE,prompt=FALSE}
sprayMeans
```

## Activity
Write a for loop that calculates the cumulative sum of the vector `c(7,2,22,5,33,8)`. 

If you don't know, the cumulative sum is a sequence of the partial sums of a given sequence. For example, the cumulative sum of the vector `c(1,2,3)` is (1,3,6).

```{r cumsum_solution,include=FALSE}
v <- c(7,2,22,5,33,8)
cumulative <- vector(mode = "numeric",length=length(v))
for (i in 1:length(v)) {
  cumulative[i] <- sum(v[1:i])
}
identical(cumulative,cumsum(v))
```

## `while` loops
- `while` loops are like a hybrid of an `if else` and `for` loops
    - They repeat the same code block over and over again like `for` loop
    - But they only repeat the block `if` some condition is met
```{r_simple while}
k <- 1
while(k < 5) {
  print(k)
  k <- k + 1
}
```

## `while` loops
- `while` loops are approriate to write when you don't know *a priori* how many iterations you will need
    - e.g. How many samples from the normal must you take before getting a value more than 3 SD's away?
```{r normaldeviates}
r <- rnorm(1)
samples <- 1
while(r < 3 && r > -3) {
    samples <- samples+1
    r <- rnorm(1)
}
print(paste("Success only took",samples,"samples"))
```

## `while` loops
A good example of looping indefinitely until a condition occurs is the brute-force method of calculating how many trials you will need to acheive a specific level of experimental power

```{r powerWhileLoop,prompt=F}
power <- 0; alt_p <- .75; null_p <- .5; n <- 1;
while (power < .8) {
  crit_n <- qbinom(.025, size=n, prob = null_p, lower.tail=FALSE)
  power <- pbinom(crit_n, size=n, prob = alt_p, lower.tail = FALSE)
  n <- n+1
}
n-1 # Number of trials needed
```

## loop pro-tips
- Loops are slow, and as a consequence, you should avoid large loops. 
- If you must use one (and sometimes you must), **ALWAYS PREALLOCATE THE FINAL DATA STRUCTURES YOU WANT**
  - Remember in our InsectSprays means example, we made a place holder data frame the exact size and shape we wanted, and then filled it in
  
```{r ref.label="sideeffects_step2", eval=F,prompt=FALSE}
```

## `stop`
- You can halt the execution of your program using a `stop` statement
    - You probably don't need this just yet in your programming lives, but its good to know it exists
- One good use for this is in combination with `if` statements, to detect states that are indicative of serious problems, and throw an error 

```{r stopinthenameoflove,eval=F,prompt=FALSE}
x <- somenumbers
if (sd(x) < 0) {
  stop("Standard Deviation cannot be negative!")
}
```
