---
title: 'Text Processing Basics'
author: "William Hopper"
date: "`r Sys.Date()`"
---

## Text Processing Basics
R is primarily a numeric computing environment, but  has a wide array of tools to process textual data as well.

By textual data, I mean data stored in character vectors, e.g., 
```r
c("I am textual data!")
```

The individual elements of character vectors are formally called *character constants* but are colloquially referred to to as *strings*.

Today, we'll cover some basic functions for manipulating strings and character vectors, introduce tools to perform pattern matching with strings, and introduce the `stringr` package.

## Elements vs. Characters
There is an important distinction between the number of elements in a character vector, and the number of characters in a single string.

For example, the character vector `c("I", "Love", "You")` has 3 elements ("I","Love", and "You"), but its individual strings have 1, 4, and 3 characters, respectively.

Use the `length()` function to determine the number of elements in a vector, and use the `nchar()` function to determine the number of characters in each string of the vector.

```{r nchar}
length(c("I", "Love", "You"))
nchar(c("I", "Love", "You"))
```

## Changing Cases
If you want to change *all* the characters in a vector to lower or uppercase characters, we can use the `tolower()` and `toupper()` functions.

```{r}
tolower("Good MORNING!")
toupper("Good MORNING!")
```

This is useful when you want to compare 2 vectors for equality, without the case of the letters getting in the way of the comparison.

```{r}
tolower("SEND HELP!") == "send help!"
```

## Combining Character Vectors
The `paste` function is used to combine two or more character vectors into one vector.

```{r}
paste("You're", "a", "wizard", "Harry", "!")# 5 Individual strings, same as c() here
```

This function is quite useful when you have a string represented as a variable, and you would like to place together with other text (e.g., to make a sentence).

```{r}
paste("The current date and time is: ", date())
```

## Combining Character Vectors
If the vectors have more than one element, and each vector has the same length, their elements will be pasted together pairwise.

```{r}
paste(c("Good","Light","Right"),
      c("Bad", "Dark", "Wrong"))
```

You can also choose a string that will separate the concatenated elements by using the `sep` argument (the default is a space).

```{r}
paste(c("Good","Light","Right"),
      c("Bad", "Dark", "Wrong"),
      sep= " vs. ")
```

## Combining Character Vectors
If the vectors have more than one element, but each vector *does not* have the same length, R will do some weird recycling of each vectors elements...

```{r}
paste(c("A","B"),
      c("C","D","E"), 
      c("V","W","X","Y","Z"))
```
`c("A","B")` gets extended to `c("A","B","A","B","A")` and `c("C","D","E")` gets extended out to be `c("C","D","E","C","D")`. This is probably not what you want, so be careful.

## Combining Character Vectors
`paste()` can also combine the many elements of a character vector into a single element, by supplying a value for the `collapse` argument.

```{r}
paste(c("1","2","3"), collapse = " + ")
```

You can also use this together with multiple character vectors, but be careful, because the vectors elements are concatenated before the final product is collapsed!

```{r}
paste(c("1","2","3"), c("4","5","6"),
      collapse = " + ")
```

"1" and "4", "2" and "5", and "3" and "6" are concatenated with a space, then "1 4", "2 5" and "3 6" are collapsed with a " + ".

## Splitting vectors apart
If we want to extract some of the characters within a string, and we know the specific *positions* of the characters within a string we want to extract (e.g., the first through the fifth characters), we can use the `substr()` function.

`substr()` has 3 arguments:

- `x`: the vector whose elements should have some characters extracted
- `start`: the *position* of the first character to be extracted in each element
- `stop`: the *position* of the last character to be extracted in each element

```{r}
substr("abcdefg", start=2, stop=4)
```

## Splitting vectors apart
`substr` is a vectorized function, so we can extract sub-strings from each string in character vectors with more than 1 elements. You can also make `start` and `stop` the same value to extract just one character.

```{r}
substr(c("Hit","Eye","Low","Lamb","Oval"), start=1, stop=1)
```
We can also extract *different* substrings from each element in the character vector by supplying longer vectors to `start` and `stop`.

```{r}
substr(c("Pill","xzzy","haa!"), start=c(1,2,3), stop=c(2,3,4))
```

Here, we extract the first and second characters from "Pill", the second and third characters from "xzzy", and the third and fourth characters from "haa!".

## Patern Matching
What if you want to extract part of a string, but don't know for sure where the part you want will be located? Or what if you want to count the number of times a specific word occurs in some text, but the text isn't broken up by individual words already?

If you know how to describe the *pattern* of text you're looking for, then you can use *regular expressions* to search for strings and substrings that match your pattern of interest.

Regular expressions (or regexes, for short) are a double edged sword: they are extremely powerful and useful, but can easily [summon Codethulu](https://thedailywtf.com/articles/The_Call_of_Codethulhu). We'll just touch on some basics today, and leave the dangerous stuff for another class.

## grep
R's `grep()` function (an acronym for " <b>g</b>lobal <b>r</b>egular <b>e</b>xpression <b>p</b>rint") is commonly used to search for matches to a regular expression pattern within the elements of a character vector.

`grep()`has 2 required arguments:

1. `pattern`: A character string containing a regular expression
2. `x`: A character vector where matches to `pattern` are searched for

By default, `grep()` returns the locations of the elements in the character vector that match the regex. 

But, we can have it return the actual character strings that were matched by setting the `value` argument to `TRUE`.

## Literal Character Patterns
The simplest regular expressions include only literal characters, e.g., a, b, c, 1, 2, 3, etc.

For example, the regex "cat" matches the strings, "category", "catnip", and "concatenate", because the characters "c", then "a", then "t", literally appear in each one.

But, it wouldn't match "cart", because the "r" interrupts the pattern.

```{r}
grep("cat", c("category", "catnip", "concatenate", "cart"), value=TRUE)
```

## Character Classes
We can also specify a *set* of possible characters to match. If *any* of the characters in the set occur in the string, it is considered a match.

These sets are called "character classes", and are denoted by surrounding the characters you want to match with square brackets.

So, if we wanted to find a match to a "c" *or* an "a", *or* a "t" occurring *anywhere* in a string, we could use the regexes `[cat]`, or `[act]`, or `[tac]`, etc., order doesn't matter in a character class.

```{r}
grep("[tac]", c("cat", "dog", "cart", "bar"), value=TRUE)
```

## Ranges of Characters
If you want to specify a large range of characters for your set, the `-` symbol provides a convenient shorthand you can use to define *ranges* of characters.

For example, if I wanted to match any of the characters "a", "b", "c", "d", "e", "f", "g", "h", "1", "2", "3", "4", "5", or "6", I could type them all out, or use the following character ranges:

```{r}
grep("[abcdefgh12345]",c("funky","town","is","#1"),value=TRUE)
grep("[a-h1-6]",c("funky","town","is","#1"),value=TRUE)

```

