---
title: "Reshaping Datasets with tidyr"
author: "William Hopper"
date: "`r Sys.Date()`"
output:
  ioslides_presentation:
    css: ["../theme/css/custom.css", "../theme/css/tidyCellHighlighting.css"]
---

```{r setup,echo=FALSE,message=FALSE,warning=FALSE}
library(pander)
library(tidyr)
library(dplyr)
library(whoppeR)
knitr::opts_chunk$set("warning"=FALSE, "message"=FALSE, cache = TRUE)
```

## Reshaping with `tidyr`
The tools from the `tidyr` package help you re-arrange tabular data stored in data frames, with an emphasis on helping you *normalize* datasets.

There are 4 main functions for re-arranging data in the `tidyr` package:

1. `gather()`: transforms data from wide(r) to long(er)
2. `spread()`: transforms data from long(er) to wide(r)
3. `separate()`: splits a single column into multiple columns
4. `unite()`: combines multiple columns into a single column

## The temperature data
We'll demonstrate how to use each function with a small dataset, so its easy to see what changes get made. So, install `tidyr`, and load the [temperature dataset](../data/temperature.csv).

```{r, eval=FALSE, cache=FALSE}
install.packages("tidyr")
library(tidyr)
temperature <- read.csv("https://wjhopper.github.io/psych640/data/temperature.csv")
temperature
```

```{r import_temp_data, echo=FALSE, cache=FALSE}
temperature <- read.csv("../data/temperature.csv")
tmp_initial <- temperature
```
```{r display_temp_data, echo=FALSE}
pander(temperature, "High Temperatures (Farenheight)")
```

## The temperature data
```{r, ref.label="display_temp_data", echo=FALSE}
pander(temperature, "High Temperatures (Farenheight)")
```

The `temperature` dataset has measurements of 3 variables: **City**, **Date**, and **Temperature**.

But the layout of the dataset is non-normalized, because values of the **Date** variable (Feb_19_2017, Feb_18_2017, and Feb_17_2017) are used as unique variables themselves.

Instead, the dates *should* be in the cells of the table as observed values of the **Date** variable.

## The `gather()` function 
Let's use the `gather` function to move the dates in the column headers down into the table under a new variable **date**, and move the actual temperature measurements into a new variable called **temp**.

`gather` takes 4 required arguments:

1. `data`: the name of the data frame to re-arrange
2. `key`: the name of the to-be-created variable that will hold values currently in use as variables
3. `value`: the name of the to-be-created variable that will hold the values from the variables to be gathered.
4. `...`: A bare list of columns to gather (or columns *not* to gather)


## The `gather()` function 
Let's make this more specific to our situation.

```{r,eval=FALSE}
gather(data = temperature,
       key = "date",
       value = "temp",
       -city)
```

- The `key` column will be named "date", because it will hold the dates that were previously used as individual variables
- The `value` column will be named "temp" because the values in each of the Feb_19_2017, Feb_18_2017, and Feb_17_2017 variables were our actual temperature measurements.
- We want to gather values from all the columns *except* the city column, so we can write `-city`. Alternatively, we could write `Feb_19_2017, Feb_18_2017, and Feb_17_2017`.

## The `gather()` function {#gather  .sothis}
So this:
```r
temperature <- gather(data = temperature, key = "date", value = "temp", -city)
```
turns this:
```{r, echo=FALSE}
pander(temperature)
```

into this:

```{r, echo=FALSE}
temperature <- gather(data = temperature, key = "date", value = "temp", -city)
pander(temperature)
```


## The `gather()` function {#gather2}
Critically, the temperature observations are still matched up with the same date!

```{r, echo=FALSE}
pander(tmp_initial) 
```

```{r, echo=FALSE}
pander(temperature)
```


## The `spread()` function
The `spread` function is the complement to the `gather` function, so we can use it to undo this transformation. 

Its job is to takes values from an existing variable in the dataset and "spread" them out, creating a new variable for each unique value in the old variable.

The `spread` function takes 3 required arguments:

1. `data`: the name of the data frame to re-arrange
2. `key`: the name of the existing variable whose *N* unique values will be "spread out" into *N* new variables
3. `value`: the name of the existing variable whose values will be used as observations in the *N* to-be-created variables.

## The `spread()` function
So to undo the wider-to-longer normalization we just performed, we would use this function call:

```{r,eval=FALSE}
spread(data = temperature,
       key = "date",
       value = "temp")
```

- the `key` argument is the "date" column, because the "date" column holds the values we want to be new variables
- the `value` argument is the "temp" column, because we want the temperature measurements to be observations in the 3 date variables

We don't have to specify the names of the new variable we're creating, because they are going to come from the values in the "date" variable.

## The `spread()` function {#spread  .sothis}
So this:
```r
spread(data = temperature, key = "date", value = "temp")
```
turns this:
```{r, echo=FALSE}
pander(temperature)
```

into this:

```{r, echo=FALSE}
pander(spread(data = temperature, key = "date", value = "temp"))
```


## Common `spread()` problems
A common issue encountered with the `spread` function is the somewhat mysterious error "<code style="color: red;">Duplicate identifiers for rows ... </code>".

Basically, this means is that the rows of your data frame are not uniquely identifiable, and `spread` can't figure out which observations belong in each row when it reshapes that data. Rather than potentially corrupt your data, it gives up instead.

In general, the solution is to give each row a unique ID number like so:

```{r eval = FALSE} 
temperature$id <- 1:nrow(temperature)
```

## Common `spread()` problems {.errormsg}
For example, I can't spread the `X` variable in this data frame into new `A` and `B` variables.
```{r notunique,echo=FALSE}
set.seed(10)
d <- data.frame(X = c("A","A","B","B"),
                Y = c(1, 1, 1, 2),
                Z = sample(10:200, 4))
pander(d)
```

```{r trytospread, error=TRUE}
spread(d, key=X, value=Z)
```
The problem is that when trying to form the `A` column, it's ambigious whether the 106 or the 68 should go in the first row and get paired up with 90, because both 106 and 68 have a `Y` value of 1.


## Common `spread()` problems
If we add an observation identifier within A and B, we can reshape the data.
```{r adduniqueID,echo=FALSE}
d$id <- rep(1:2, 2)
pander(d)
```

```{r show_spread, eval=FALSE}
spread(d, key=X, value=Z)
```
```{r actuallyspread, echo=FALSE}
pander(spread(d, key=X, value=Z)) 
```

## The `separate()` function
The `separate` function is used to transform one variable in your dataset into many variables in your dataset.
This is useful when multiple variables have been concatenated into a single variable, and you would like to examine these variables separately.

For example, in our normalized `temperature` dataset, the `date` variable is actually a combination of a month, day of month, and year variable, concatenated together.

```{r, echo=FALSE}
pander(temperature)
```

## The `separate()` function
Let's separate this amalgamated "date" variable into 3 different variables, one for the month, day, and year.

The `separate` function has 3 required arguments, and a 4^th^ optional argument you almost always want to specify.

1. `data`: the name of the data frame to re-arrange
2. `col`: the bare name of the variable to be split-up
3. `into`: a character vector of names for the to-be-created variables
4. `sep`: the delimiter defining how to split existing values into multiple new values. 

By default, `separate` will split the values in the `col` variable on any non-alphanumeric character (e.g., `_`,  `-`, `.`, `:`, etc.). You can have it split on a character string, a numeric index (e.g., every 3^rd^ character), or using a regular expression.

## The `separate()` function
In our case, we would use the separate function like so:
```{r separate, eval=FALSE}
separate(data = temperature,
         col = date,
         into = c("month", "day", "year"),
         sep = "_")
```

- We want to separate the `col`umn named `date` *into* 3 new variables named "month", "day", and "year"
- We are going to find the values that should go into the month, day, and year variables by splitting up the values in the `date` variable at every underscore character.

## The `separate()` function {#separate .sothis}
So this:
```r
temperature <- separate(temperature, col=date, into=c("month", "day", "year"), sep = "_")
```

turns this:
```{r, echo=FALSE}
pander(prettifyTable(temperature, rows = 3))
```

into this:
```{r, echo=FALSE}
temperature <- separate(temperature, col=date, into=c("month", "day", "year"), sep = "_")
pander(prettifyTable(temperature, rows = 3))
```

## The `unite()` function
The `unite` function is the complement to the separate function. As you can imagine, its job is to combine multiple variables into a single variable. 

The `unite` function has 3 required arguments, and 2 other arguments you often want to specify:

1. `data`: the name of the data frame to re-arrange
2. `col`: the bare name of the to-be-created amalgam variable
3. `...` : a list of bare columns whose values are to be combined row-wise and placed inside the new variable `col`
4. `sep`: the separator to use between combined values (defaults to `"_"`).
5. `remove`: a logical value indicating whether or not to remove columns from the data frame after combining them (defaults to `TRUE`).

## The `unite()` function {.sothis}
So to undo our separation of the `date` variable, we could use this:
```r
unite(temperature, col=date, month, day, year)
```

to turn this:
```{r, echo=FALSE}
pander(prettifyTable(temperature, rows = 3))
```

back into this:
```{r, echo=FALSE}
pander(prettifyTable(unite(temperature,
                           col=date,
                           month, day, year),
                     rows = 3))
```

## Resources & More
For more information about the *theory* behind the `tidyr` package, check out the [package vignette](https://cran.r-project.org/web/packages/tidyr/vignettes/tidy-data.html) and the [tidy data paper](http://vita.had.co.nz/papers/tidy-data.html).

You may have noticed that the API (i.e., standard syntax) for `tidyr` functions is not unlike that of the functions in `dplyr`. This is not a coincidence; both packages have the same core authors and are part of the "[tidyverse](https://blog.rstudio.org/2016/09/15/tidyverse-1-0-0/)" collection of R packages.

However, `tidyr` is *not* the only choice for reshaping data. The `reshape2` package is also quite useful. Its focus is on reshaping more broadly (e.g., from one type of data structure to another, reshaping through aggregation, etc.). If you need a more general reshaping tool, start with the following `reshape2` tutorials:

- [My reshape2 lab](https://wjhopper.github.io/psych640/labs/reshape.html)
- [A brief but good intro to reshape2](http://seananderson.ca/2013/10/19/reshape.html)
- [Hadley Wickham's intro to reshape2](http://had.co.nz/reshape/introduction.pdf)


## Activity
Normalize the `iris` dataset using tidyr functions.

```{r print_iris, echo=FALSE}
pander(prettifyTable(iris, 6))
```

Go!
```{r activity, include=FALSE}
x <- iris %>%
  gather(measurement, size, -Species) %>%
  separate(measurement, c("structure", "dimension")) %>%
  group_by(Species) %>%
  mutate(id = 1:n()) %>%
  ungroup() %>%
  arrange(Species, id) %>%
  spread(Species, size)
```